---
title: "Iteration and Looping"
output: html_notebook
---

## Conditionals in R
Conditionals are statements that evaluate a logical statement and then make a choice. They are a simple programming concept that can build in a lot of power and flexibility to your R scripting. The most straight-forward of the conditional statements are the simple __if__ and __if/else__.

Remember that the symbol for logical equivalence is the double equal sign __==__.
```{r}
# Simple if statement
day <- "Tuesday"
if (day=="Tuesday") {
  print("today")
}

# Simple if/else example
if (day=="Monday") {
  print("yesterday")
} else {
  print("not yesterday")
}

# more complicated if/else example
if (day=="Monday") {
  print("yesterday")
} else if (day=="Tuesday") {
  print("today")
} else {
  print("not yesterday")
}
```
Anything that evaluates to __TRUE__ or __FALSE__ can be put into a conditional statement. You can also use the logical operators __&__(and), __|__(or), or __!__(not) in your conditional statements.
```{r}
library(lubridate)  # This is an amazing package!
julian.date <- yday(today())
yr <- year(today())
if (julian.date >= 65) {
  print("It's after March 6th")
}

if (julian.date != 65) {
  print("It's not March 6th")
}

if (julian.date == 65 & yr == 2019) {
  print("March 6th, 2019")
}

```


## Iteration and looping
Iteration and looping in R is useful for when you need to do the same operation over a set of objects, datasets, or observations, or when you want to run some simulations. You can easily set up iteration in R over multiple variables (as long as you can keep your head around that they're doing!). The basic iterator in R is the __for()__ function.

Check out this example of the basic for loop syntax.
```{r}
for (i in 1:10) {  # Basic structure 
  print(i)        # i is the "iterator". It's a counter in this case.
}
```
One important thing to note about iterating using __for__ in R is that R takes care of moving through all of the iterations on it's own. In the example above, you don't need to tick __i__ up by one each time. (This is not always the case with other programming languages!).

Another iterator in R is __while__. While is useful for when you need to repeat something until you meet some sort of condition - i.e., you don't know how many times you want to repeat ahead of time. Examples of this would be running a simulation until it stabilizes or reaches some threshold, or recursive parsing of a document for text patterns. _Be careful with while(), though, because it's easy to get into never-ending (infinite) loops._

```{r}
# simple while example to replicate a for loop
i <- 0
while (i < 10) {
  print(i)
  i<-i+1  # Need to count up i by one each time in this case.
}

now <- Sys.time()
while (Sys.time() < now+10) {
  print("waiting...")
}
print("It's now 10 seconds later!")

```


You can also iterate over lists of objects. This becomes really powerful if you need to do the same thing over a bunch of datasets.
```{r}
for (month in c("Jan","Feb","Mar","Apr","May")) {
  print(month)
}
```

Use different R functions to create sets of objects to iterate over
```{r}
for (rand.num in rnorm(12)) { # Iterate over a set of 12 random numbers
  print(paste("The Random number is: ",rand.num))
}
```

Grab a list of files from a directory and iterate over those
```{r}
data.path <- "C:\\Users\\jakal\\OneDrive - University of Idaho\\Documents\\GitHub\\REM504-DataScience\\Practice_Datasets"
files <- list.files(data.path,pattern="*.csv")
out.data <- c()
for (file in files) {
  print(file)
  out.data[[file]]<-read.csv(paste(data.path,file,sep="\\"))
}
```

For simulations, you can use loops to build your results
```{r}
out.data <- data.frame() # set up an empty data frame
for (i in 1:10) {
  var.rand <- rnorm(1)
  var.now <- date()
  var.other <- "Jason"
  out.row <- data.frame(i=i, rand=var.rand, now=var.now, me=var.other) # Assemble your new data
  out.data <- rbind(out.data,out.row) # Append it to your out dataframe
}
glimpse(out.data)
```


You can also nest for loops
```{r}
out.data <- data.frame() # set up an empty data frame
for (x in seq(-3,3,by=0.1)) {  # Note use of seq() here.
  for (y in seq(-3,3,by=0.1)) {
    z <- exp(-((x-0)^2/2+(y-0)^2/2)) ## 2D gaussian formula with mean zero and variance of 1
    out.row <- data.frame(x=x, y=y, z=z)
    out.data <- rbind(out.data,out.row)
  }
}
glimpse(out.data)
ggplot(data=out.data,aes(x=x,y=y,fill=z))+geom_raster()
```

